---
title: "Estimating and Testing IV models in R"
author: "Elad Guttman"
output:
  xaringan::moon_reader:
    seal: false
    css: ["middlebury", "middlebury-fonts" , "libs/mytemp.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'

---

class: title-slide-section-blue, middle, center

# Estimating and Testing IV models in R

##Elad Guttman

---
class: large

# Outline

1. [A Reminder](#reminder)

2. [One endogenous variable](#one)

3. [Multiple endogenous variables](#multiple)


---
class: title-slide-section-blue, middle, center
name: reminder

#A Reminder

---
class: large

# The Model

- The general model you discussed in class:

$$y = X\beta + u = X_1\beta_1 + X_2\beta_2 + u$$

- where:

  - $x_1$ are $k_1$ exogenous variables
  
  - $x_2$ are $k_2$ endogenous variables
  
  - $K = k_1 + k_2$

- $z$ is a $1 \times L$ vector of: 
  
  - $k_1$ exogenous variables
  
  - $m$ instruments
  
  - $L = k_1 + m$

---
class: large

# The Identification Assumptions

1. **2SLS.1:** $E(z'u) = 0$ - all the variables in $z$ are uncorrelated with $u$

2. **2SLS.2:** 

  a. $rank(E(z'z)) = L$ - no multicollinearity
  
  b. $rank(E(z'x)) = K$ - The instruments are correlated with $X$, and the model is at least just-identified ( $L \geq K$ )
  
---
class: large

# The 2SLS Estimator

1. **First stage:** for each endogenous variable $j$ we estimate:
$$x_j = Z\delta + \epsilon$$ using OLS.

2. **Second stage:** then we replace $x_2$ with $\hat{x}_2$ (i.e., the predicted values from the first stage) and estimate:
$$y = X_1\beta_1 + \hat{X}_2\beta_2 + u$$ using OLS (again). Under assumptions **2SLS.1** and **2SLS.2**, this is a valid procedure for obtaining a consistent estimate for $\beta$.

---
class: title-slide-section-blue, middle, center
name: one

# One endogenous variable

---
class: large

# Data

For this exercise, we'll use the following dataset:
```{r, include=F}
help_console <- function(topic, format=c("text", "html", "latex", "Rd"),
                         lines=NULL, before=NULL, after=NULL) {  
  format=match.arg(format)
  if (!is.character(topic)) topic <- deparse(substitute(topic))
  helpfile = utils:::.getHelpFile(help(topic))
  hs <- capture.output(switch(format, 
                              text=tools:::Rd2txt(helpfile),
                              html=tools:::Rd2HTML(helpfile),
                              latex=tools:::Rd2latex(helpfile),
                              Rd=tools:::prepare_Rd(helpfile)
                              )
                      )
  if(!is.null(lines)) hs <- hs[lines]
  hs <- c(before, hs, after)
  cat(hs, sep="\n")
  invisible(hs)
}

```

```{r, message=F, warning=F}
library(wooldridge)
data("card")
#function to print the content from an R help file. 
#You don't need that, just use ?card instead
help_console(card, "text", 5:8) 
```

---
class: large

#Data

We'll pay special attention to the following variables:
```{r}
help_console(card, "text", c(22:27, 30:33, 70:71))
```

---
class: large

# Estimation

The familiar `lfe` package provides flexiable syntax for estimating IV models. Here are some examples: 

```{r, message=F, warning=F}
library(lfe)
library(stargazer)
library(tidyverse)

card = card %>%
  select(lwage, nearc2, nearc4, fatheduc, motheduc, educ) %>%
  drop_na() #caution: in general, this is a bad practice!

#one endogenous variable, two exogenous variables, and one instrument
twosls_with_exog = felm(lwage ~ fatheduc + motheduc | 0 | (educ ~ nearc4), data = card)

#one endogenous variable and one instrument, no exogenous variables
twosls_no_exog = felm(lwage ~ 1 | 0 | (educ ~ nearc4), data = card)

#we can easily look at the first stage results:
stage1_with_exog = twosls_with_exog$stage1
stage1_no_exog = twosls_no_exog$stage1
```

---

## First Stage

```{r}
stargazer(stage1_no_exog, stage1_with_exog, dep.var.labels = "educ",
          type = "text", keep.stat = c("n"))

```

---
## Second Stage

```{r}
stargazer(twosls_no_exog, twosls_with_exog, type = "text", keep.stat = c("n"))
```

---
## Standard Errors

```{r}
#As usual, the felm function also calculates robust SE:
stage1_no_exog$robustvcv
twosls_no_exog$robustvcv

summary(twosls_no_exog, robust = T)$coefficients
```

---
# Manual Estimation

Alternatively, we can estimate IV models manually by estimating two OLS models: 

```{r}
manual_stage1_with_exog = lm(educ ~ fatheduc + motheduc + nearc4, data = card) 

card$`educ(fit)` = predict(manual_stage1_with_exog)
manual_twosls_with_exog = lm(lwage ~ fatheduc + motheduc + `educ(fit)`, data = card)
```

---
## First Stage

```{r}
stargazer(manual_stage1_with_exog, stage1_with_exog ,dep.var.labels.include = F, dep.var.caption = "Model:",
          type = "text", keep.stat = c("n"))
```

---

## Second Stage

.pull-left[
```{r, echo=F}
stargazer(manual_twosls_with_exog, twosls_with_exog ,dep.var.labels.include = F, dep.var.caption = "Model:",
          type = "text", keep.stat = c("n"))
```
]

--

.pull-right[
  .red[.large[Keep in mind that the right standard errors <br>
  aren't the standard errors of the second stage!]]
]

---

# Testing for weak instruments

- Remember that weak instrument can cause the 2SLS estimator to be biased towards OLS. 

- We can test that by testing the null that the coefficients on the excluded instruments are 0.

- This is a simple Wald / F-test.

```{r}
r = 0
R = matrix(c(0, 0, 0, 1), nrow = 1)
waldtest(stage1_with_exog, R, r)
```

---

# Testing for weak instruments

`felm` calculates this F-statistic by default (and also the robust version):
```{r}
stage1_with_exog$iv1fstat
stage1_with_exog$rob.iv1fstat
```

---
class: title-slide-section-blue, middle, center
name: multiple

# Multiple endogenous variables

---

# Estimation

Very easy with the `felm` function:

```{r}
#genetate squared education,
#so that we'll have two endogenous variables:
card = card %>%
  mutate(educ2 = educ^2)

twosls = felm(lwage ~ fatheduc + motheduc | 0 | (educ | educ2 ~ nearc4 + nearc2), data = card)

stage1 = twosls$stage1
```

---

## First Stage

```{r}
summary(stage1, lhs = "educ")$coefficients
```


```{r}
summary(stage1, lhs = "educ2")$coefficients
```

---

## Second Stage

.pull-left[
```{r, echo=F}
stargazer(twosls, type = "text", keep.stat = c("n"))
```
]

--

.pull-right[
.red[.large[In the following slides we'll see that
            this estimation suffers from a weak IV problem,
            so do not try to interpret these results]]
]
---

# Testing for weak instruments

```{r}
stage1$iv1fstat
```

--

<br><br>
.red[.large[What is this test? Is it enough?]]

---
class: large

# Testing for weak instruments

- This is an equation-by-equation F-test

- For each first stage equation, we test $H_0: \beta_{nearc2} = \beta_{nearc4} = 0$

- Remember the alternative: $H_1: \beta_{nearc2} \not= 0$ .red[or] $\beta_{nearc4} \not= 0$

- So a single instrument could be correlated with all the endogenous regressors,
while the rest of the instruments are not

- This is not enough

---
class: large

# Conditional F-test

The **general** idea proposed by Angrist and Pischke (2009):

- Consider a model with two endogenous variables, $x_1$ and $x_2$ and a valid vector of instruments $z$

- Let's look at the following model: $y = \beta_1x_1 + \beta_2\hat{x_2} + u*$
  
  - $\hat{x_2}$ is exogenous since $\hat{x_2} = P_zx_2$
  
  - $x_1$ is endogenous and (potentially) correlated with $\hat{x_2}$
  
---
class: large

# Conditional F-test
  
- Now let's regress $x_1$ on $\hat{x_2}$ and keep the residuals $M_{\hat{x_2}}x_1$

- Let's look at the following model: $y = \beta_1M_{\hat{x_2}}x_1 + \epsilon$

  - $M_{\hat{x_2}}x_1$ is endogenous, but not correlated with $\hat{x_2}$
  
  - This is the short version of the previous equation
  
  - Applying the 2SLS procedure yields a consistent estimate for $\beta_1$
  
- So now we have only one endogenous variable, and we can use the usual F-test in the first stage: $M_{\hat{x_2}}x_1 = \kappa Z + \epsilon*$

--

- .red[Note: This is just the general idea. See Sanderson and Windmeijer (2016) for the formal test]

---
class: large

# Conditional F-test

`condfstat` from the `lfe` package performs that test:
```{r}
condfstat(twosls)
```